### 2012-06-05 paper/chemokine.tex
Intermediate changes

The adaptive immune response induced during an acute viral infection is a complex web of defense mechanisms able to control all but the most virulent of infections.   A complete understanding of the evolution of this response will be instrumental in the design of antiviral prophylaxis, therapy and vaccine interventions.   The effector arm of the immune response operates, however, under strict constraints of time and the geography of body structure during the dynamic spread of replicating viruses.  Elegant studies have outlined the role of local dendritic cells in detecting invasion and delivering antigen-specific cargoes to the lymph nodes [1-3].  In situ imaging has provided temporal and spatial details on the communications between dendritic cells and recirculating precursor or memory T and B cells within the lymph node architecture [4-6].  The specificity, activation profile and functional capabilities of resulting effector lymphocytes are well described [7-9].  In contrast to the induction phase of the response, how activated lymphocytes emanating from regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10], has been less well studied.

The immune response to influenza A virus (IAV) is arguably the most well-studied with respect to viral specificity, individual immunocyte components, whole animal models, and mathematical modeling.  Animal models of IAV include the non-human primate, ferret, and mouse [11].   Although not a natural host of influenza viruses, the inbred mouse model has provided a detailed description of viral and immunocyte dynamics.  Mathematical models have used the mouse model data to simulate the evolution of the immune response and its impact on viral kinetics [12-16].  These mathematical models have approached such questions as the relative roles of antibody and cytotoxic T lymphocytes (CTL), number of CTL required, regulation of antigen presentation, and comparing drug intervention strategies.  Such whole-response modeling has the potential advantage of selecting the most promising strategies of vaccination and therapy for subsequent expensive animal and human trials.   Each of these models, however, must make assumptions about the efficiency of each functional component in the model.   Math modeling can also inform us on the behavior of each component.

Here we approach the more focused problems of time and physical constraints on the lymphocyte diaspora.  How do very small foci of infected tissue attract limited numbers of activated CTL released into the systemic vascular system servicing non-infected tissue orders of magnitude larger than the target infection?  A host of chemokines secreted by infected cells, with signal amplification by immigrating inflammatory cells, are critical in directing inflammatory and immune cells to sites of inflammation [17-19].  While each chemokine has been studied in detail for its synthesis, receptor specificity and function in knockout models, the spatial and temporal details of chemokine function and efficiency have not been explicitly modeled in a virus-infected whole animal environment.

Using an agent-based model (ABM) designed to test the temporal and spatial demands during an acute influenza infection in the lung, we examined the homing of CTL to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.   Viral secretion dynamics and detailed chemokine secretion patterns by infected human epithelial cells in vitro [20] provided the data representing local behaviors at the alveolar level.  Data on CTL production and recirculation and the resolution of viral infections was obtained from the literature on mouse infections.   We excluded the role of immigrating inflammatory cells in order to focus on the contribution of two chemokines induced during infection with three influenza viruses, including a seasonal H1N1 virus, the pandemic H1N1 virus, and a virulent avian H5N1 virus.  In our modeling environment we observed the markedly different efficiencies of control for the three influenza viruses, differing effectiveness of the two chemokines, and the importance of the velocity of CTL search within tissue.

@@

>>>>

The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.  Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes [1-3], temporal and spatial details of precursor T and B cell activation within the lymph node architecture [4-6], and functional capabilities of resulting effector lymphocytes [7-9].  In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10].

Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics [12-16].  Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.    Models using differential equations to predict temporal events do not account for spatial characteristics of cell behavior, and must make assumptions about the efficiency of each functional component in the model.

Here we focus on the time and physical constraints of the lymphocyte diaspora.  How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control (17a), after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?  A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection [17-19].  While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.

Using an agent-based model (ABM) designed to test the spatial demands on the effector phase, we examined the homing of activated cytotoxic T lymphocytes (CTL) to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.   Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains [20], combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model. In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.

@@

\subsection*{Model Validation}

The ABM produces infection dynamics consistent with \textit{in vivo} observations.  We observe FOI diameters on the order of 1mm in seasonal H1N1 simulation.  The T-cell response at day five post-infection rapidly decreases the number of infected cells over the first 24 hours.

Scaling to humans?

\subsection*{ABM Advantages}

The use of an ABM has advantages over a spatially homogeneous ODE model.  An ODE model assumes that any virus particle is capable of infecting any healthy cell.  Figure \ref{fig:cycells} shows that this is clearly not the case.  In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.  ODE models account for this discrepancy by lowering rates of infection, but this simply results in improper parameter values.

The spatial nature of the simulation also reveals challenges to T-cell detection rates that are not apparent in ODE models.  T-cells find infected cells by climbing the chemokine gradient.  This results in T-cell clustering at local maxima of chemokine concentration.  Thus, T-cells do not increase in effectiveness as the number of T-cells increases.  Furthermore, infected cells grow more disperse as the size of the infection grows.  Because T-cells cluster together they become incapable of covering the increasing plaque size.  This allows for the infection to recover and accelerate once its size crosses a certain threshold.  This is another effect that is not seen in an ODE model.

\subsection*{Caveats and Limitations}

In both H1N1 strains the T-cell resposne fails to clear the infection by day 10.  In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.  We suggest that this is due to the absence of several innate immune agents that were not included in the model.  We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.  Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.  We see that the viral production rate is a major factor in the strength of the infection.

\subsection*{Future Work}
